---
title: "What Helps Students Read Better? Key Findings from International Data"
author: "Changqi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: false
    fig_width: 8
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
```

## The Big Picture

We analyzed reading scores from over **thousands of students across dozens of countries** using the PISA 2018 international assessment. Our goal: understand what factors—at the student and school level—help kids become better readers.

---

## Key Findings

```{r load-data}
df <- read_csv("../data/processed/analysis_data.csv")
```

### 1. Internet Access is Linked to Higher Reading Scores

```{r internet-plot}
df %>%
  filter(!is.na(internet)) %>%
  group_by(internet) %>%
  summarise(avg_reading = mean(reading_score, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = internet, y = avg_reading, fill = internet)) +
  geom_col(alpha = 0.85, show.legend = FALSE, width = 0.5) +
  geom_text(aes(label = round(avg_reading, 0)), vjust = -0.5, size = 6, fontface = "bold") +
  scale_fill_manual(values = c("no" = "#e67e22", "yes" = "#27ae60")) +
  ylim(0, 600) +
  labs(
    title = "Students with internet access score higher in reading",
    x = "Internet Access at Home",
    y = "Average Reading Score"
  ) +
  theme_minimal(base_size = 14)
```

> Students with internet access at home score significantly higher on reading assessments, even after accounting for family income.

### 2. Girls Outperform Boys in Reading — Everywhere

```{r gender-plot}
gender_gap <- df %>%
  filter(!is.na(gender)) %>%
  group_by(country, gender) %>%
  summarise(avg = mean(reading_score, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = gender, values_from = avg) %>%
  mutate(gap = female - male) %>%
  arrange(desc(gap)) %>%
  head(15)

gender_gap %>%
  ggplot(aes(x = reorder(country, gap), y = gap)) +
  geom_col(fill = "#9b59b6", alpha = 0.85) +
  geom_text(aes(label = round(gap, 0)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  labs(
    title = "Girls outscore boys in reading across all countries",
    subtitle = "Top 15 countries by gender gap",
    x = NULL,
    y = "Score Difference (Girls − Boys)"
  ) +
  theme_minimal(base_size = 13)
```

### 3. Family Income Matters — But Schools Can Help

```{r ses-plot}
df %>%
  filter(!is.na(ses_index), !is.na(school_avg_ses)) %>%
  mutate(school_ses_group = if_else(school_avg_ses > median(school_avg_ses, na.rm = TRUE),
                                     "Higher-Resource Schools",
                                     "Lower-Resource Schools")) %>%
  sample_n(min(3000, nrow(.))) %>%
  ggplot(aes(x = ses_index, y = reading_score, color = school_ses_group)) +
  geom_point(alpha = 0.12, size = 0.8) +
  geom_smooth(method = "lm", linewidth = 1.3) +
  scale_color_manual(values = c("Higher-Resource Schools" = "#27ae60",
                                 "Lower-Resource Schools" = "#e74c3c")) +
  labs(
    title = "The SES-achievement gap varies by school context",
    subtitle = "Higher-resource schools may help narrow the gap",
    x = "Student Socioeconomic Status",
    y = "Reading Score",
    color = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

> While family income is a strong predictor of reading scores, the school a student attends can make a difference. Students in well-resourced schools tend to perform better regardless of individual background.

---

## What This Means

- **Technology access** (like internet at home) is associated with better reading outcomes.
- **School environment** matters: well-resourced schools can help close achievement gaps.
- **Targeted interventions**—like providing engaging, level-appropriate reading content—could make a real difference for students in under-resourced settings.

---

*Data source: PISA 2018 International Assessment | Analysis conducted in R*
