---
title: "How School Resources Shape Student Reading Achievement: A Multilevel Analysis of PISA 2018 Data"
author: "Changqi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
    fig_width: 9
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(lme4)
library(lmerTest)
library(performance)
library(broom.mixed)
```

# Introduction

Reading literacy is foundational to academic success and lifelong learning. While individual factors like socioeconomic status (SES) are known predictors of reading achievement, less is understood about how **school-level resources**—such as internet access, staffing, and funding structures—contribute to or mitigate achievement gaps.

This study uses data from the **Programme for International Student Assessment (PISA) 2018** to investigate:

1. How much of the variation in reading scores occurs between schools and countries versus within?
2. Which student-level and school-level factors predict reading achievement?
3. Does school socioeconomic composition moderate the relationship between individual SES and reading?

# Data & Methods

## Data Source

We use the [learningtower](https://kevinwang09.github.io/learningtower/) R package, which provides cleaned subsets of PISA data. PISA is an international assessment administered every three years to 15-year-old students across 80+ countries, with 2018 being the most recent reading-focused cycle.

## Variables

**Student-level:**

- `reading_score` — PISA reading assessment score (outcome)
- `gender_female` — binary indicator (1 = female)
- `ses_centered` — socioeconomic index, grand-mean centered
- `has_internet` — binary indicator for home internet access

**School-level:**

- `school_avg_ses` — average SES of students in the school
- `is_public` — binary indicator (1 = public school)
- `staff_shortage` — index of staff shortage severity

```{r load-data}
df <- read_csv("../data/processed/analysis_data.csv")

model_df <- df %>%
  select(reading_score, gender_female, ses_centered, ses_index,
         has_internet, school_avg_ses, is_public, staff_shortage,
         school_id, country) %>%
  drop_na()

cat("N =", nrow(model_df), "students across",
    n_distinct(model_df$country), "countries and",
    n_distinct(model_df$school_id), "schools")
```

## Analytical Approach

We fit a series of **multilevel (hierarchical) linear models** with students nested in schools nested in countries using the `lme4` package. We build models progressively:

1. **Null model** — Unconditional to compute ICC
2. **Student-level model** — Add individual predictors
3. **School-level model** — Add school characteristics
4. **Interaction model** — Cross-level interaction (SES × school SES)

# Results

## Exploratory Analysis

```{r eda-distribution, fig.cap="Distribution of PISA 2018 reading scores."}
ggplot(model_df, aes(x = reading_score)) +
  geom_histogram(bins = 50, fill = "#3498db", color = "white", alpha = 0.8) +
  labs(title = "Distribution of Reading Scores", x = "Reading Score", y = "Count") +
  theme_minimal(base_size = 13)
```

```{r eda-ses, fig.cap="SES is strongly associated with reading achievement."}
model_df %>%
  sample_n(min(2000, nrow(model_df))) %>%
  ggplot(aes(x = ses_centered, y = reading_score)) +
  geom_point(alpha = 0.15, size = 0.8, color = "#2c3e50") +
  geom_smooth(method = "lm", color = "#e74c3c", linewidth = 1.2) +
  labs(title = "SES vs Reading Score", x = "SES (centered)", y = "Reading Score") +
  theme_minimal(base_size = 13)
```

## Model 1: Null Model (ICC)

```{r model-null}
model_null <- lmer(reading_score ~ 1 + (1 | country/school_id), data = model_df)
icc_result <- icc(model_null)
print(icc_result)
```

The ICC tells us what proportion of variance in reading scores lies between schools and countries. A high ICC justifies the use of multilevel modeling.

## Model 2: Student-Level Predictors

```{r model-student}
model_student <- lmer(
  reading_score ~ gender_female + ses_centered + has_internet +
    (1 | country/school_id),
  data = model_df
)
summary(model_student)
```

## Model 3: School-Level Predictors

```{r model-school}
model_school <- lmer(
  reading_score ~ gender_female + ses_centered + has_internet +
    school_avg_ses + is_public + staff_shortage +
    (1 | country/school_id),
  data = model_df
)
summary(model_school)
```

## Model 4: Cross-Level Interaction

```{r model-interaction}
model_interaction <- lmer(
  reading_score ~ gender_female + ses_centered * school_avg_ses +
    has_internet + is_public + staff_shortage +
    (1 | country/school_id),
  data = model_df
)
summary(model_interaction)
```

## Model Comparison

```{r model-comparison}
anova(model_null, model_student, model_school, model_interaction)
```

```{r r-squared}
r2_values <- tibble(
  Model = c("Null", "Student", "School", "Interaction"),
  R2_marginal = c(
    r2(model_null)$R2_marginal,
    r2(model_student)$R2_marginal,
    r2(model_school)$R2_marginal,
    r2(model_interaction)$R2_marginal
  ),
  R2_conditional = c(
    r2(model_null)$R2_conditional,
    r2(model_student)$R2_conditional,
    r2(model_school)$R2_conditional,
    r2(model_interaction)$R2_conditional
  )
)
knitr::kable(r2_values, digits = 4, caption = "R-squared for each model")
```

## Coefficient Plot

```{r coef-plot, fig.cap="Fixed effects with 95% confidence intervals from the final multilevel model."}
tidy_results <- tidy(model_interaction, effects = "fixed", conf.int = TRUE)

tidy_results %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = reorder(term, estimate), y = estimate)) +
  geom_point(size = 3, color = "#2c3e50") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#e74c3c") +
  coord_flip() +
  labs(title = "Fixed Effects from Final Model", x = NULL, y = "Estimate (95% CI)") +
  theme_minimal(base_size = 13)
```

# Discussion

This analysis demonstrates how multilevel modeling can partition variance in student reading scores across individual, school, and country levels. Key findings:

- **SES matters at both levels**: Individual SES and school-average SES are both significant predictors.
- **Internet access** is associated with higher reading scores even after controlling for SES.
- **The cross-level interaction** reveals whether the SES–achievement relationship varies by school context.

These findings are directly relevant to educational technology platforms (like Newsela) that aim to close achievement gaps by providing differentiated, accessible content to students across diverse schools.

# Technical Notes

- All analyses conducted in R `r R.version.string` using `lme4`, `lmerTest`, and `tidyverse`
- Data from PISA 2018 via the `learningtower` R package
- SQL queries for data extraction available in `sql/queries.sql`
